Lattices of acyclic pipe dreams

### 2023-05-25 acyclicFacets/draft/latticePipeDreams.tex
minor corrections and comments/questions
minor corrections

Given a pipe dream $P$, we write $i\to j$ whenever $i$ has an elbow weakly northwest of $j$. \cref{lem:rectangle} gives us that $i\to j$ implies $i\contactLess{P} j$. 
When $P$ is not acyclic, we can find a cycle ${\bf C}=(i_1, i_2, \ldots, i_m)$ of the contact graph, where $i_1 \to i_2\to\cdots\to i_m\to i_1$. 
We say that $\bf{C}$ is a \defn{reduced cycle} if it is not possible to remove any pipe from $\bf C$ and still have a cycle. 
Given a reduced cycle $\bf C$, we can also assume that $i_1 = \min\{i_1, \dots, i_m\}$.

@@

For the converse, assume we have a pipe dream $P$ that is not acyclic. The contact graph must contain a reduced cycle ${\bf C}=(i, j_1, \ldots, j_m)$ with $i<j_k$ for all $k \in [m]$. 
If $m=1$, then $i$ and $j_1$ must cross , $i\to j_1$ implies there is an elbow of $i$ southwest of the crossing and $j_1\to i$ implies there is an elbow of $j$ northeast of the crossing. \cref{lem:notDominant} implies $\omega$ is not dominant. We consider the case where $m>1$.
For $i\to j_1$, there is an elbow of~$i$ at $(a,b)$ weakly northwest of an elbow of~$j_1$ at $(c_1,d_1)$ an elbow.
Similarly, for $j_m\to i$, there is an elbow of~$j_m$ at $(a_m,b_m)$ weakly northwest of an elbow of~$i$ at~$(c,d)$.
Since the cycle $C$ is reduced, we must have that $(a_m,b_m)$ is not northwest $(c_1,d_1)$. That is $c_1<a_m$ or $d_1<b_m$.

@@

We are grateful to Florent Hivert for various discussions. We also thank Martin Rubey for his comments which led us to think about Theorem~\ref{prob:nuAcyclicProperty} and
we are very grateful to Lucas Gagnon who shared with us an independent proof of the converse direction of Theorem~\ref{prob:nuAcyclicProperty}, this allowed us to simplify grately our argument.

>>>>

Given a pipe dream $P$, we write $i\to j$ whenever $i$ has an elbow weakly northwest of $j$. \cref{lem:rectangle} gives us that $i\to j$ implies $i\contactLess{P} j$. 
When $P$ is not acyclic, we can find a cycle ${\bf C}=(i_1, i_2, \ldots, i_m)$ of the contact graph, where $i_1 \to i_2\to\cdots\to i_m\to i_1$. 
We say that $\bf{C}$ is a \defn{reduced cycle} if it is not possible to remove any pipe from $\bf C$ and still have a cycle. 
Given a reduced cycle $\bf C$, we can also assume that $i_1 = \min\{i_1, \dots, i_m\}$.
\vincent{Maybe I would put this cycle story inside the proof.}

@@

For the backward direction, assume we have a pipe dream $P$ that is not acyclic. The contact graph must contain a reduced cycle ${\bf C}=(i, j_1, \ldots, j_m)$ with $i<j_k$ for all $k \in [m]$. 
If $m=1$, then $i$ and $j_1$ must cross , $i\to j_1$ implies there is an elbow of $i$ southwest of the crossing and $j_1\to i$ implies there is an elbow of $j$ northeast of the crossing. \cref{lem:notDominant} implies $\omega$ is not dominant. We consider the case where $m>1$.
For $i\to j_1$, there is an elbow of~$i$ at $(a,b)$ weakly northwest of an elbow of~$j_1$ at $(c_1,d_1)$ an elbow.
Similarly, for $j_m\to i$, there is an elbow of~$j_m$ at $(a_m,b_m)$ weakly northwest of an elbow of~$i$ at~$(c,d)$.
Since the cycle $C$ is reduced, we must have that $(a_m,b_m)$ is not northwest $(c_1,d_1)$. That is $c_1<a_m$ or $d_1<b_m$.

@@

We are grateful to Florent Hivert for various discussions. We also thank Martin Rubey for his comments which led us to think about~\cref{prob:nuAcyclicProperty}.
Finally, we are grateful to Lucas Gagnon who shared with us an independent proof of the backward direction of \cref{prob:nuAcyclicProperty}, allowing us to greatly simplify our argument.
